Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system of plausible mechanism

### 2024-08-06 index.qmd
section:background,dags; process:write

<!-- 4. But which approach are we going to use? -->
Nevertheless, the structural approach offers an additional key advantage over the potential outcomes approach: it enables the graphical representation of any system through directed acyclic graphs [DAG, @Gross_et_al_2018; @Neal_2020]. DAGs serve as heuristics, effectively conveying the assumed causal structure of a system. A heuristic is a strategy that simplifies information to make decisions more quickly, efficiently, and sometimes more accurately than complex methods [@Chow_2015]. Consequently, DAGs do not represent detailed statistical models but allow researchers to deduce which statistical models can provide valid causal inferences, assuming the causal structure depicted in the DAG is accurate [@McElreath_2020].

@@

<!-- 3. What are the advantages of DAGs? -->
DAGs offer two key advantages for modeling causality. Firstly, they represent causal relations in a nonparametric and fully interactive manner. This generality implies that feasible causal analysis strategies usually do not require specifying the type of data or the nature of the functional dependence among variables [@Morgan_et_al_2014]. Secondly, regardless of complexity, DAGs can represent various causal structures using only five fundamental building blocks [@Neal_2020; @McElreath_2020]. These building blocks can be depicted in three ways: the magnified representation, the standard representation, and the structural causal model form [SCM, @Morgan_et_al_2014].

@@

<!-- 3.3 the SCM -->
Lastly, the right panels of @fig-dags_scms depict the SCM representation of the five fundamental building blocks. SCMs are formal mathematical models that represent causal relationships within a system or population [@Hitchcock_2024]. An SCM is defined by the set of endogenous variables $V$, the set of exogenous variables $E$, and the set of functions $F=\{f_{X},f_{Z},f_{Y}\}$ [@Pearl_2009; @Neal_2020]. These functions, referred to as structural equations, specify each endogenous variable as a nonparametric function of other variables. Moreover, SCMs use the symbol '$:=$' to indicate the variables' asymmetrical causal dependence and the symbol '$\dsep$' to represent *d-separation*, which roughly equates to the concept of variable independence. The concepts of d-separation and causal (in)dependence are explored in the following section.

@@

<!-- 4. But what does the building blocks mean? -->
A careful examination of @fig-dags_scms highlights the assumptions underlying these building blocks. Figures [-@fig-mdag_bb1], [-@fig-sdag_bb1], and SCM [-@fig-scm_bb1] depict two unconnected nodes, representing a scenario where variables $X$ and $Y$ are not causally related. Figures [-@fig-mdag_bb2], [-@fig-sdag_bb2], and SCM [-@fig-scm_bb2] illustrate two connected nodes, showing a scenario where a *parent* node $X$ exerts a causal influence on a *child* node $Y$. Consequently, $Y$ is considered a *descendant* of $X$. Figures [-@fig-mdag_bb3], [-@fig-sdag_bb3], and SCM [-@fig-scm_bb3] depict a *chain* or *pipe*, where $X$ influences $Z$, and $Z$ influences $Y$. In this configuration, $X$ is a parent node of $Z$, and $Z$ is a parent node of $Y$. Furthermore, the DAGs show that $X$ is an *ancestor* of $Y$ and that $Z$ fully *mediates* the relationship between these variables. Figures [-@fig-mdag_bb4], [-@fig-sdag_bb4], and SCM [-@fig-scm_bb4] illustrate a *fork*, where variables $X$ and $Y$ are both influenced by $Z$. Here, $Z$ is a parent node of $X$ and $Y$. Finally, Figures [-@fig-mdag_bb5], [-@fig-sdag_bb5], SCM [-@fig-scm_bb5] depict a *collider*, also known as *inmorality*, where variables $X$ and $Y$ are concurrent causes of $Z$. In this configuration, $X$ and $Y$ are not causally related to each other but both influence $Z$. Additionally, in all SCMs, the errors are assumed to be mutually independent of each other and from all other variables in the graph, as evidenced by the pairwise relations $e_{X} \dsep e_{Y}$, $e_{X} \dsep e_{Z}$, and $e_{Z} \dsep e_{Y}$.

<!-- 5. A motivating example -->
Given the heuristic nature of DAGs, a motivating example can clarify how to use the five fundamental building blocks to construct the causal structure of a system. Consider a research problem where the causal effect of a variable $T$ on an outcome $Y$ needs investigation. The problem suggests that a variable $X$ potentially influences both $T$ and $Y$. Beyond these relationships, the problem does not specify any further variables of interest. Such scenarios are not hard to imagine. For instance, $T$ might represent different treatments that could affect the recovery from cancer $Y$, while $X$ could denote the patient's age. Similarly, in the context of a DCJ study like the one described by @Boonen_et_al_2020, $T$ could represent the duration of a child's cochlear implant use, which might influence the child's overall speech quality $Y$, with $X$ indicating the child's hearing status. [(not a bad example, but I prefer one using writing skills)]{style="color:blue;"}

<!-- 6. DAG for the motivating example -->
@fig-example illustrates the plausible causal structure of the motivating example. A detailed examination of Figures [-@fig-mdag_example1], [-@fig-sdag_example1], and SCM [-@fig-scm_example1] reveals the presence of at least four of the five fundamental building blocks. The figures display multiple descendants, as indicated by pairwise relations such as $X \rightarrow T$, $X \rightarrow Y$, and $T \rightarrow Y$. Additionally, the figures features multiple pairs of unconnected nodes, evident from the relations $e_{T} \dsep e_{X}$, $e_{T} \dsep e_{Y}$, and $e_{X} \dsep e_{Y}$. Finally, the figures illustrate a fork with $X \rightarrow \{T,Y\}$, as well as colliders such as $\{X,e_{T}\} \rightarrow T$ and $\{X,T,e_{Y}\} \rightarrow Y$.

@@

<!-- \begin{aligned} -->
<!--     Y & := \; f_{Y}( X,Z,e_{Y} ) \\ -->
<!--     X & := \; f_{X}( e_{X} ) \\ -->
<!--     Z & := \; f_{Z}( e_{Z} ) \\ -->
<!--     e_{X} & \dsep e_{Z} \\ -->
<!--     e_{X} & \dsep e_{Y} \\ -->
<!--     e_{Z} & \dsep e_{Y} -->
<!-- \end{aligned} -->

<!-- these are non-parametrical structural equations -->

>>>>

<!-- 4. But which approach are we going to use? -->
However, the structural approach offers an additional key advantage over the potential outcomes approach: it enables the graphical representation of any system through directed acyclic graphs [DAG, @Gross_et_al_2018; @Neal_2020]. DAGs serve as heuristics, effectively conveying the assumed causal structure of a system. A heuristic is a strategy that simplifies information to make decisions more quickly, efficiently, and sometimes more accurately than complex methods [@Chow_2015]. Consequently, DAGs do not represent detailed statistical models but allow researchers to deduce which statistical models can provide valid causal inferences, assuming the causal structure depicted in the DAGs are accurate [@McElreath_2020].

@@

<!-- 3. What are the advantages of DAGs? -->
DAGs offer two key advantages for modeling causality. Firstly, DAGs represent causal relations in a nonparametric and fully interactive manner. This feature implies that feasible causal analysis strategies usually do not require specifying the type of data or the nature of the functional dependence among variables [@Morgan_et_al_2014]. Secondly, regardless of complexity, DAGs can represent various causal structures using only five fundamental building blocks [@Neal_2020; @McElreath_2020]. These building blocks can be depicted in three ways: the magnified representation, the standard representation, and the structural causal model form [SCM, @Morgan_et_al_2014].

@@

<!-- 3.3 the SCM -->
Lastly, the right panels of @fig-dags_scms depict the SCM form of the five fundamental building blocks. SCMs are formal mathematical models that represent causal relationships within a system or population [@Hitchcock_2024]. An SCM is defined by a set of endogenous variables $V$, a set of exogenous variables $E$, and a set of functions $F=\{f_{X},f_{Z},f_{Y}\}$ [@Pearl_2009; @Neal_2020]. These functions, referred to as structural equations, specify each endogenous variable as nonparametric functions of other variables. Moreover, SCMs use the symbol '$:=$' to indicate the variables' asymmetrical causal dependence and the symbol '$\dsep$' to represent *d-separation*, which roughly equates to the concept of variable independence. The concepts of d-separation and causal (in)dependence are explored in the following section.

@@

<!-- 4. But what does the building blocks mean? -->
A careful examination of @fig-dags_scms highlights the assumptions underlying these building blocks. Figures [-@fig-mdag_bb1], [-@fig-sdag_bb1], and SCM [-@fig-scm_bb1] depict two unconnected nodes, representing a scenario where variables $X$ and $Y$ are not causally related. Figures [-@fig-mdag_bb2], [-@fig-sdag_bb2], and SCM [-@fig-scm_bb2] illustrate two connected nodes, showing a scenario where a *parent* node $X$ exerts a causal influence on a *child* node $Y$. Consequently, $Y$ is considered a *descendant* of $X$. Figures [-@fig-mdag_bb3], [-@fig-sdag_bb3], and SCM [-@fig-scm_bb3] depict a *chain* or *pipe*, where $X$ influences $Z$, and $Z$ influences $Y$. In this configuration, $X$ is a parent node of $Z$, and $Z$ is a parent node of $Y$. Furthermore, the DAGs show that $X$ is an *ancestor* of $Y$ and that $Z$ fully *mediates* the relationship between these variables. Figures [-@fig-mdag_bb4], [-@fig-sdag_bb4], and SCM [-@fig-scm_bb4] illustrate a *fork*, where variables $X$ and $Y$ are both influenced by $Z$. Here, $Z$ is a parent node of $X$ and $Y$. Finally, Figures [-@fig-mdag_bb5], [-@fig-sdag_bb5], SCM [-@fig-scm_bb5] depict a *collider*, also known as *inmorality*, where variables $X$ and $Y$ are concurrent causes of $Z$. In this configuration, $X$ and $Y$ are not causally related to each other but both influence $Z$. Additionally, in all SCMs, the errors are assumed to be mutually independent of each other and of all other variables in the graph, as evidenced by the pairwise relations $e_{X} \dsep e_{Y}$, $e_{X} \dsep e_{Z}$, and $e_{Z} \dsep e_{Y}$.

<!-- 5. A motivating example -->
Given the heuristic nature of DAGs, a motivating example can help clarify how to use the five fundamental building blocks to construct the causal structure of a system. Consider a research problem where the causal effect of a variable $T$ on an outcome $Y$ needs investigation. The problem suggests that a variable $X$ potentially influences both $T$ and $Y$. Beyond these relationships, the problem does not specify any further variables of interest. Such scenarios are not hard to imagine. For instance, $T$ might represent different treatments that could affect the recovery from cancer $Y$, while $X$ could denote the patient's age. Similarly, in the context of a DCJ study like the one described by @Boonen_et_al_2020, $T$ could represent the duration of a child's cochlear implant use, which might influence the child's overall speech quality $Y$, with $X$ indicating the child's hearing status. [(not a bad example, but I prefer one using writing skills)]{style="color:blue;"}

<!-- 6. DAG for the motivating example -->
@fig-example illustrates the plausible causal structure of the motivating example. A detailed examination of Figures [-@fig-mdag_example1], [-@fig-sdag_example1], and SCM [-@fig-scm_example1] reveals the presence of at least four of the five fundamental building blocks. The figures display multiple descendants, as indicated by pairwise relations such as $X \rightarrow T$, $X \rightarrow Y$, and $T \rightarrow Y$. Additionally, the figures features multiple pairs of unconnected nodes, evident from the relations $e_{T} \dsep e_{X}$, $e_{T} \dsep e_{Y}$, and $e_{X} \dsep e_{Y}$. Finally, the figures illustrate the fork $X \rightarrow \{T,Y\}$, and two colliders with $\{X,e_{T}\} \rightarrow T$ and $\{X,T,e_{Y}\} \rightarrow Y$.

@@
